\propertitle

### 2019-12-19 maxNt_v6.tex
update

\pdfoutput=1
%% Author: PGL  Porta Mana
%% Created: 2015-05-01T20:53:34+0200
%% Last-Updated: 2019-12-16T16:01:08+0100
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newif\ifarxiv
\arxivfalse
\ifarxiv\pdfmapfile{+classico.map}\fi
\newif\ifafour
\afourfalse% true = A4, false = A5
\newif\iftypodisclaim % typographical disclaim on the side
\typodisclaimtrue
\newcommand*{\memfontfamily}{zplx}
\newcommand*{\memfontpack}{newpxtext}
\documentclass[\ifafour a4paper,12pt,\else a5paper,10pt,\fi%extrafontsizes,%
onecolumn,oneside,article,%french,italian,german,swedish,latin,
british%
]{memoir}
\newcommand*{\updated}{\today}
\newcommand*{\firstdraft}{4 November 2015}
\newcommand*{\firstpublished}{***}
\newcommand*{\propertitle}{Inferring the total activity of a large neuronal population\\ from a small sample}
\newcommand*{\pdftitle}{\propertitle}
\newcommand*{\headtitle}{Inferences about large neuronal populations}
\newcommand*{\pdfauthor}{P.G.L.  Porta Mana, Y. Roudi, V. Rostami, E. Torre}
\newcommand*{\headauthor}{Porta Mana \etal}
\newcommand*{\reporthead}{}% Report number

@@

As a first step we need to translate vague terms like
\enquote{cooperativity}, \enquote{interaction}, and similar into a more
precise, operational \citep{bridgman1927_r1958} notion. Here we propose a
notion that
\begin{enumerate*}[label=(\roman*)]
\item seems closely related to those terms as used in the
  literature; \item is mathematically connected with maximum-entropy
  distributions; \item reveals important consequences of sampling.
\end{enumerate*}
This notion is \emph{informational sufficiency}
\citep{kullbacketal1951}[\sect~4.5]{bernardoetal1994}[\chap~8 and
\sect~14.2]{jaynes1994_r2003}. Its meaning and the three points above will
now be illustrated; please see the end of this section for references to
in-depth discussions and proofs of technical results.

Our belief distribution about the frequencies of the neuronal activities,
or about the activity in a new time bin, is in principle conditional on all
experimental data and statistics we have. This distribution, however, can
be such that it remains unchanged if we discard part of the data or
statistics -- for example, the third- and higher-order empirical moments --
but it does change if we discard more. This means that the discarded
statistics are \emph{informationally irrelevant}: their knowledge does not
change our beliefs. The remaining statistics -- in the example, first and
second empirical moments -- are \emph{minimally informationally
  sufficient}. The qualification \enquote{minimally} will be omitted for
brevity.
% This informational quasi-sufficiency is interesting because
% it may give clues about population organization and dynamics.

Now let us turn the informational sufficiency of a subset of statistics and
the informational sufficiency of another subset into two hypotheses. We can
then ask: which hypothesis is more probable, given the data? equivalently,
what is the ratio of their probabilities? Translating our question about
\enquote{cooperativity} into a hypothesis-testing problem about sufficiency
gives it a well-defined answer determined by the
probability-calculus. % The ratio of
% the two probabilities gives us a natural measure of the
% For example, do
% the data say that the first two empirical moments are probably enough to
% determine our beliefs? or is it more probable that the first four are
% needed instead?

But it does more: through a mathematical theorem, it leads to the automatic
appearance of maximum-entropy distributions: \emph{if a probability
  distribution for a repetitive phenomenon has some sufficient statistics,
  then it is a mixture of distributions of maximum-entropy form having
  those statistics as constraints, and vice versa}. We give detailed
references about this theorem at the end of this section.
% Thus, maximum-entropy distributions automatically appear (as
% approximations) from the probability-calculus when we test hypotheses
% about informational sufficiency.
Its most important consequence is that the probabilities for the two
hypotheses are approximately proportional, by Bayes's theorem, to the
probabilities for all the data according to the two maximum-entropy
distributions constrained on the subsets of statistics (provided we have
equal pre-data beliefs about the two hypotheses). We shall shortly explain
and perform this calculation for the first data set.

@@

An important fact must be taken into account, however: \emph{informational
  sufficiency is not preserved under sampling}. If a probability
distribution has some sufficient statistics, then its sample marginals
\emph{cannot} have the same sufficient statistics, and vice versa
\citep[\sect~3.1]{portamanaetal2015}; except for trivial cases such as
uniform probability distributions. % This generally holds also among
% different marginals: if a statistics is sufficient for some marginal, then
% it will not be sufficient for another marginal.
This impossibility is known in another guise in statistical mechanics: if a
system is described by a Gibbs state, its subsystems cannot be perfectly
described by Gibbs states \citep[\eg][and references
therein]{maesetal1999}. % Mathematically this impossibility comes from the
% Pitman-Koopman theorem mentioned above, and translates into the general
% impossibility of solving a system of independent equations with more
% equations than unknowns .

This fact is important for interpreting the results of our application. For
example, if means and pairwise correlations seem informationally sufficient
for a particular sample from a brain area, then they may well not be
sufficient for the larger population of neurons constituting that area, and
vice versa. So if we are interested in the \enquote{cooperativity} or
\enquote{interaction} of the neurons in a brain area, it is unreliable to
use a maximum-entropy distribution constructed only for a sample
thereof. % \citep[contrast this with some literature, \eg][]{martignonetal1995,bohteetal2000,schneidmanetal2006,shlensetal2006,barreiroetal2010,ganmoretal2011,granotatedgietal2013}
The model presented here avoids this problem because the maximum-entropy
method is applied directly to the larger population, not to the sample
alone.

@@

For the model at the sample level, that is, with $N=n$, the weights of
evidence are
\begin{equation}
  \label{eq:diff_suff_n}
  \begin{split}
  \ydi_{n}(M_{4},M_{2}) &= 280\;\nat = 1\,220\;\hart,
\\    \ydi_{n}(M_{5},M_{4}) &= 3.3\;\nat = 1.4\;\hart.
\end{split}
\end{equation}
The data are 1\,220 orders of magnitude more probable conditional on four
moments than conditional on two, and $10^{1.4}=25$ times more probable
conditional on five moments than on four moments.

The conclusions about \enquote{cooperativity} or \enquote{interaction} that
we reach by applying the maximum-entropy method to the larger population,
as proposed here, are therefore different from those applying it at the
sample level -- and also visually clearer. The plot in 
\fig~\ref{fig:plots_2_4mom_a}, showing the two clearly different
distributions constrained by two and four moments, should be compared with
the plot for the distributions obtained at the sample level under the same
constraints, shown in \fig~\ref{fig:comparisons_marginals} as
\textcolor{myred}{red circles} and \textcolor{myblue}{blue squares}. The
measured frequencies (black triangles) are also shown. It is necessary to
use a logarithmic scale to see the differences, which appear mainly in the
tail.
\begin{figure}[!p]
\centering
% \includegraphics[width=\linewidth]{_scripts/comparison_full.pdf}%
% \\[3em]%
\includegraphics[width=0.95\linewidth]{_scripts/trunct_comparison_marginals.pdf}%
\caption{Marginals for the normalized total activity of the sample,
  obtained from the maximum-entropy distributions with $N=10\,000$ from
  \fig~\ref{fig:plots_2_4mom_a}, and from maximum-entropy distributions
  calculated at the sample level, that is with $N=n=65$, with two and four
  moments in either case.}
\label{fig:comparisons_marginals}
\end{figure}
% comparison_ME_sample_full_levels.nb

@@

It is also possible to use the measure~\eqref{eq:measure_suffic} and its
probabilistic interpretation~\eqref{eq:bayes_factor} to compare the probability of
the data $\yf$ conditional on the hypothesis $(M_{4},N)$ of four-moment
sufficiency at the larger-population level, $N=10\,000$, versus the
hypothesis $(M_{4}, n)$ of four-moment sufficiency at the sample level,
$N=n=65$. We obtain
\begin{equation}
  \label{eq:diff_suff_n}
  \ydi[(M_{4}, N), (M_{4}, n)] = 39\;\nat = 17\;\hart.
\end{equation}
% that is, the data are 17 orders of magnitude more probable under the first
% hypothesis than under the second. 
This result is also shown in
\fig~\ref{fig:comparisons_marginals}, where we see that the four-moment
distribution constructed at the larger-population level
(\textcolor{myblue}{blue filled squares}) fits the measured frequency
distribution (black triangles) more closely than the corresponding
sample-level distribution (\textcolor{myblue}{blue empty squares}).

>>>>

\pdfoutput=1
%% Author: PGL  Porta Mana
%% Created: 2015-05-01T20:53:34+0200
%% Last-Updated: 2019-12-19T20:55:39+0100
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newif\ifarxiv
\arxivfalse
\ifarxiv\pdfmapfile{+classico.map}\fi
\newif\ifafour
\afourfalse% true = A4, false = A5
\newif\iftypodisclaim % typographical disclaim on the side
\typodisclaimtrue
\newcommand*{\memfontfamily}{zplx}
\newcommand*{\memfontpack}{newpxtext}
\documentclass[\ifafour a4paper,12pt,\else a5paper,10pt,\fi%extrafontsizes,%
onecolumn,oneside,article,%french,italian,german,swedish,latin,
british%
]{memoir}
\newcommand*{\updated}{\today}
\newcommand*{\firstdraft}{4 November 2015}
\newcommand*{\firstpublished}{***}
\newcommand*{\propertitle}{Inferring the total activity of a large neuronal population\\ from a small sample}
\newcommand*{\pdftitle}{\propertitle}
\newcommand*{\headtitle}{Inferences about large neuronal populations}
\newcommand*{\pdfauthor}{P.G.L.  Porta Mana, Y. Roudi, V. Rostami, E. Torre}
\newcommand*{\headauthor}{Porta Mana \etal}
\newcommand*{\reporthead}{}% Report number

@@

As a first step we need to translate vague terms like
\enquote{cooperativity}, \enquote{interaction}, and similar into a more
precise, operational \citep[\chap~I]{bridgman1927_r1958} notion. Here we
propose the notion of \emph{informational sufficiency} (also called
\enquote{predictive sufficiency})
\citep{kullbacketal1951}[\sect~4.5]{bernardoetal1994}[\chap~8 and
\sect~14.2]{jaynes1994_r2003}, because
\begin{enumerate*}[label=(\roman*)]
\item it seems closely related to those terms as used in the
  literature; \item it is mathematically connected with maximum-entropy
  distributions; \item it reveals important consequences of sampling.
\end{enumerate*}
Its meaning and the three points above will now be illustrated; please see
the end of this section for references to in-depth discussions and proofs
of technical results.

Our belief distribution about the frequencies of the neuronal activities,
or about the activity in a new time bin, is in principle conditional on all
experimental data and statistics we have. This distribution, however, can
be such that it remains unchanged if we discard part of the data or
statistics -- for example, the third- and higher-order empirical moments. % --
% but it does change if we discard more.
This means that the discarded statistics are \emph{informationally
  irrelevant}: knowing them does not change our beliefs. The remaining
statistics -- in the example, first and second empirical moments -- are
\emph{informationally sufficient}.%  The qualification \enquote{minimally}
% will be omitted for brevity.
% This informational quasi-sufficiency is interesting because
% it may give clues about population organization and dynamics.

Now let us turn the informational sufficiency of a subset of statistics and
the informational sufficiency of another subset into two hypotheses. We can
then ask: which hypothesis is more probable, given the data? equivalently,
what is the ratio of their probabilities? By translating our question about
\enquote{cooperativity} into a hypothesis-testing problem about sufficiency,
we can find a well-defined answer determined by the
probability-calculus. % The ratio of
% the two probabilities gives us a natural measure of the
% For example, do
% the data say that the first two empirical moments are probably enough to
% determine our beliefs? or is it more probable that the first four are
% needed instead?

But such a translation does more: through a mathematical theorem, it leads
to the automatic appearance of maximum-entropy distributions: \emph{if a
  probability distribution for a repetitive phenomenon has some sufficient
  statistics, then it is a mixture of distributions of maximum-entropy form
  having those statistics as constraints, and vice versa}. See the end of
this section for detailed references about this theorem.
% Thus, maximum-entropy distributions automatically appear (as
% approximations) from the probability-calculus when we test hypotheses
% about informational sufficiency.
Its most important consequence is that, by Bayes's formula, the probability
that some statistics are informationally sufficient can be approximated by
finding the maximum-entropy distribution constrained by those statistics
and then calculating the probability for the whole data under that
distribution. We shall shortly illustrate this kind of calculation for the
first data set.

@@

The theorem above makes it very natural to reinterpret studies about
cooperativity, interaction, and similar, based on maximum-entropy, as
studies about informational sufficiency. An important fact must then be
taken into account: \emph{informational sufficiency is not preserved under
  sampling}. If a probability distribution has some sufficient statistics,
then its sample marginals \emph{cannot} have the same sufficient
statistics, and vice versa \citep[\sect~3.1]{portamanaetal2015}; except for
trivial cases such as uniform probability
distributions. % This generally holds also among
% different marginals: if a statistics is sufficient for some marginal, then
% it will not be sufficient for another marginal.
This impossibility is known in another guise in statistical mechanics: if a
system is described by a Gibbs state, its subsystems cannot be perfectly
described by Gibbs states \citep[\eg][and references
therein]{maesetal1999}. % Mathematically this impossibility comes from the
% Pitman-Koopman theorem mentioned above, and translates into the general
% impossibility of solving a system of independent equations with more
% equations than unknowns .

This fact affects the interpretation of the results obtained from
maximum-entropy methods. For example, if means and pairwise correlations
seem informationally sufficient for a particular sample from a brain area,
then they are probably not sufficient for the larger population of neurons
constituting that area, and vice versa. If we are interested in the
cooperativity of the neurons in a brain area, it is therefore unreliable to
use a maximum-entropy distribution constructed only for a sample
thereof. % \citep[contrast this with some literature, \eg][]{martignonetal1995,bohteetal2000,schneidmanetal2006,shlensetal2006,barreiroetal2010,ganmoretal2011,granotatedgietal2013}
The model presented here avoids this problem because the maximum-entropy
method is applied directly to the larger population, not to the sample
alone.

@@

For the model at the sample level, that is, with $N=n$, the weights of
evidence are
\begin{equation}
  \label{eq:diff_suff_n}
  \begin{split}
  \ydi_{n}(M_{4},M_{2}) &= 280\;\nat = 1\,220\;\hart,
\\    \ydi_{n}(M_{5},M_{4}) &= 3.3\;\nat = 1.4\;\hart.
\end{split}
\end{equation}
So the application at the sample level says that the data are 1\,220 orders
of magnitude more probable conditional on four moments than conditional on
two, and $10^{1.4}=25$ times more probable conditional on five moments than
on four moments. These results are different from the ones obtained from
the larger population, and they would be hard to deduce visually from a
look at the distributions of \fig~\ref{fig:comparisons_marginals}.

% The conclusions about \enquote{cooperativity} or \enquote{interaction} that
% we reach by applying the maximum-entropy method to the larger population,
% as proposed here, are therefore different from those applying it at the
% sample level -- and also visually clearer. The plot in 
% \fig~\ref{fig:plots_2_4mom_a}, showing the two clearly different
% distributions constrained by two and four moments, should be compared with
% the plot for the distributions obtained at the sample level under the same
% constraints, shown in \fig~\ref{fig:comparisons_marginals} as
% \textcolor{myred}{red circles} and \textcolor{myblue}{blue squares}. The
% measured frequencies (black triangles) are also shown. It is necessary to
% use a logarithmic scale to see the differences, which appear mainly in the
% tail.
\begin{figure}[!p]
\centering
% \includegraphics[width=\linewidth]{_scripts/comparison_full.pdf}%
% \\[3em]%
\includegraphics[width=0.95\linewidth]{_scripts/trunct_comparison_marginals.pdf}%
\caption{Marginals for the normalized total activity of the sample,
  obtained from the maximum-entropy distributions with $N=10\,000$ from
  \fig~\ref{fig:plots_2_4mom_a}, and from maximum-entropy distributions
  calculated at the sample level, that is with $N=n=65$, with two and four
  moments in either case.}
\label{fig:comparisons_marginals}
\end{figure}
% comparison_ME_sample_full_levels.nb

@@

It is in fact possible to use the measure~\eqref{eq:measure_suffic} and its
probabilistic interpretation~\eqref{eq:bayes_factor} to compare the
probability of the data $\yf$ conditional on the hypothesis $(M_{4},N)$ of
four-moment sufficiency at the larger-population level, $N=10\,000$, versus
the hypothesis $(M_{4}, n)$ of four-moment sufficiency at the sample level,
$N=n=65$. We obtain
\begin{equation}
  \label{eq:diff_suff_n}
  \ydi[(M_{4}, N), (M_{4}, n)] = 39\;\nat = 17\;\hart.
\end{equation}
% that is, the data are 17 orders of magnitude more probable under the first
% hypothesis than under the second. 
This result is also shown in
\fig~\ref{fig:comparisons_marginals}, where we see that the four-moment
distribution constructed at the larger-population level
(\textcolor{myblue}{blue filled squares}) fits the measured frequency
distribution (black triangles) more closely than the corresponding
sample-level distribution (\textcolor{myblue}{blue empty squares}).
